The Cost of Undisturbed Landscapes

### 2021-01-11 projects/asparagus/doc/asparagus_on_a_slide.tex
updates asparagus_on_a_slide.tex

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage[version=4]{mhchem}

@@

    \begin{frame}
        \frametitle{The cost of undisturbed landscapes}
        \framesubtitle{Some background}
        \begin{itemize}
            \item By 2030, \SI{100}{\percent}\footnote[frame]{System services and industry own consumption,
            accounting for $\sim$\SI{10}{\percent} of total electricity demand and $\sim$\SI{40}{\percent} of \ce{CO2} emissions from power
            generation are exempt} of Austria's electricity demand to be sourced from renewables
            \item Sufficient potentials for large-scale expansion only for wind and solar power
        \end{itemize}
        \medskip
        {%
            \setbeamercolor{block body}{bg=bokugreen!30, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{block}{}
                \begin{itemize}
                    \item Not disturbing landscapes by wind turbines means substituting wind turbines with PV modules (potentially open-space)
                \end{itemize}
            \end{block}
        }%
        \medskip
        \begin{itemize}
            \item Substituting wind turbines with PV modules changes system operation and investment
            \item Which cost arises due to this substitution?
            \item In our case, this is the \emph{opportunity cost of wind power}, as the cost-minimizing system relies
            (almost) entirely on wind power
        \end{itemize}
    \end{frame}

@@

        \medskip
        {%
            \setbeamercolor{block title}{bg=bokugreen, fg=white}
            \setbeamercolor{block body}{bg=bokugreen!20, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{block}{Approximation of the opportunity cost of wind turbines $OC_w$}
                %To approximate the opportunity cost of wind turbines $OC_w$, we proceed as follows:
                \begin{enumerate}
                    \item determine unconstrained level of RET that leads to minimal system cost $c_{net}$ \label{step1}
                    \item restrict the admissible expansion of the least-cost RET by a small margin, $\Delta w$, and observe
                    resulting change in system cost $\Delta c_{net}$ \label{step2}
                    \item approximate the opportunity cost of wind power by \label{step3}
                    \begin{align*}
                        OC_w = \frac{\Delta c_{net}}{\Delta w}
                    \end{align*}
                    \item repeat steps \ref{step2} and \ref{step3} till no more wind turbines can be erected
                \end{enumerate}
            \end{block}
        }%
    \end{frame}

    \begin{frame}
        \frametitle{The cost of undisturbed landscapes}
        \framesubtitle{Assumptions}
        We quantify the opportunity cost of wind power with the numerical power system model \emph{medea}, which we set up
        to resemble the Austrian and German electricity and district heating systems in 2030.
        \begin{columns}[T]
        {%
            \setbeamercolor{block title}{bg=bokugreen, fg=white}
            \setbeamercolor{block body}{bg=bokugreen!20, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{column}{7cm}
                \begin{block}{Model assumptions}
                    \begin{itemize}
                        \item perfect foresight, partial-equilibrium model of electricity and district heating sectors in
                        AT and DE
                        \item technology-rich representation of supply to meet price-inelastic demand
                        \item economic dispatch in hourly resolution plus investment
                    \end{itemize}
                \end{block}
            \end{column}

            \begin{column}{7cm}
                \begin{block}{Scenario assumptions (2030)}
                    \begin{itemize}
                        \item DE: nuclear and (partial) coal exit
                        \item DE: RET expansion as in EEG 2017
                        \item AT: must generate 79.3 TWh/a from RES
                        \begin{itemize}
                            \item + 5 TWh/a hydropower
                            \item + 1 TWh/a electricity from biomass
                        \end{itemize}
                        \item weather as in 2016 (a low wind year)
                    \end{itemize}
                \end{block}
            \end{column}
        }%
        \end{columns}
        \bigskip
        \url{https://github.com/inwe-boku/medea}
    \end{frame}

    \begin{frame}
        \frametitle{Change in annual net system cost -- Austria, 2030}
        \begin{figure}
            \includegraphics[width=0.95\textwidth]{./figures/cost_components_base_25.pdf} \label{fig:cost_breakdown}
        \end{figure}
    \end{frame}

    \begin{frame}
        \frametitle{Monthly electricity generation by source -- Austria, 2030}
        \framesubtitle{(a) maximum wind power, (b) maximum solar power}
        \begin{figure}
            \includegraphics[width=0.85\textwidth]{./figures/dispatch_AT.pdf} \label{fig:monthly_gen}
        \end{figure}
    \end{frame}

    \begin{frame}
        \frametitle{Annual opportunity cost of wind power -- Austria, 2030}
        \framesubtitle{PV capital cost: \euro 625/kWp}
        \begin{figure}
            \includegraphics[width=0.95\textwidth]{./figures/oc_base.pdf} \label{fig:opportunity_cost_base}
        \end{figure}
    \end{frame}

    \begin{frame}
        \frametitle{Sensitivity of annual opportunity cost of wind power -- Austria, 2030}
        \begin{figure}
            \includegraphics[width=0.95\textwidth]{./figures/Sensitivities.pdf} \label{fig:sensitivities}
        \end{figure}
    \end{frame}

\end{document}

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage[version=4]{mhchem}
\usepackage{qrcode}

@@

    \begin{frame}
        \frametitle{The cost of undisturbed landscapes}
        \framesubtitle{Some background}
        \begin{itemize}
            \item By 2030, \SI{100}{\percent}\footnote[frame]{System services and industry own consumption,
            accounting for $\sim$\SI{10}{\percent} of total electricity demand and $\sim$\SI{40}{\percent} of \ce{CO2} emissions from power
            generation are exempt} of Austria's electricity demand to be sourced from renewables
            \item Sufficient potentials for large-scale expansion only for wind and solar power
        \end{itemize}
        \medskip
        {%
            \setbeamercolor{block body}{bg=structure.fg!20, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{block}{}
                \begin{itemize}
                    \item Not disturbing landscapes by wind turbines means substituting wind turbines with PV modules (potentially open-space)
                \end{itemize}
            \end{block}
        }%
        \medskip
        \begin{itemize}
            \item Substituting wind turbines with PV modules changes system operation and investment
            \item Which cost arises due to this substitution?
            \item In our case, this is the \emph{opportunity cost of wind power}, as the cost-minimizing system relies
            (almost) entirely on wind power
        \end{itemize}
    \end{frame}

@@

        \medskip
        {%
            \setbeamercolor{block title}{bg=structure.fg!0, fg=structure.fg}
            \setbeamercolor{block body}{bg=structure.fg!0, fg=black}%{bg=bokugreen!20, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{block}{Approximation of the opportunity cost of wind turbines $OC_w$}
                %To approximate the opportunity cost of wind turbines $OC_w$, we proceed as follows:
                \medskip
                \begin{enumerate}
                    \item determine unconstrained level of RET that leads to minimal system cost $c_{net}$ \label{step1}
                    \item restrict the admissible expansion of the least-cost RET by a small margin, $\Delta w$, and observe
                    resulting change in system cost $\Delta c_{net}$ \label{step2}
                    \item approximate the opportunity cost of wind power by \label{step3}
                    \begin{align*}
                        OC_w = \frac{\Delta c_{net}}{\Delta w}
                    \end{align*}
                    \item repeat steps~\ref{step2} and~\ref{step3} till no more wind turbines can be erected
                \end{enumerate}
            \end{block}
        }%
    \end{frame}

    \begin{frame}
        \frametitle{The cost of undisturbed landscapes}
        \framesubtitle{Assumptions}
        {%
            \setbeamercolor{block body}{bg=structure.fg!20, fg=black}
            \setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{block}{}
                We quantify the opportunity cost of wind power with the numerical power system model \emph{medea}, which we set up
                to resemble the Austrian and German electricity and district heating systems in 2030
            \end{block}
        }%
        \begin{columns}[T]
            %{%
            %\setbeamercolor{block title}{bg=bokugreen, fg=white}
            %\setbeamercolor{block body}{bg=bokugreen!20, fg=black}
            %\setbeamertemplate{blocks}[rounded][shadow=false]
            \begin{column}{7cm}
                %\begin{block}{Model assumptions}
                \structure{Model assumptions}
                \begin{itemize}
                    \item perfect foresight, partial-equilibrium model of electricity and district heating sectors in
                    AT and DE
                    \item technology-rich representation of supply to meet price-inelastic demand
                    \item economic dispatch in hourly resolution plus investment
                \end{itemize}
                %\end{block}
            \end{column}

            \begin{column}{7cm}
                %\begin{block}{Scenario assumptions (2030)}
                \structure{Scenario assumptions (2030)}
                \begin{itemize}
                    \item DE: nuclear and (partial) coal exit
                    \item DE: RET expansion as in EEG 2017
                    \item AT: at least 74.8 TWh/a from RES
                    \begin{itemize}
                        \item + 5 TWh/a hydropower
                        \item + 1 TWh/a electricity from biomass
                    \end{itemize}
                    \item weather as in 2016 (a low wind year)
                \end{itemize}
                %\end{block}
            \end{column}
            %}%
        \end{columns}
        \bigskip
        \begin{center}
            \url{https://github.com/inwe-boku/medea}
        \end{center}
    \end{frame}

    \begin{frame}
        \frametitle{Results}
        \begin{columns}[T]
            \begin{column}{10.6 cm}
                \begin{figure}
                    \caption{Change in annual net system cost -- Austria, 2030}
                    \includegraphics[width=1.0\textwidth]{./figures/cost_components_base_25.pdf} \label{fig:cost_breakdown}
                \end{figure}
            \end{column}
            \begin{column}{5.0 cm}
                \bigskip
                \begin{itemize}
                    \item "100\%"-policy necessarily turns Austria from a net importer (2016: 7.2 TWh) into a
                    net exporter of electricity as fossil generation needed (heat, system services)
                    \item each avoided GW wind power replaced by 1.98 GWp PV
                    \item resource adequacy drives O\&M cost
                    \item lower exports also mean less fossil generation
                \end{itemize}
            \end{column}
        \end{columns}
    \end{frame}

    \begin{frame}
        \frametitle{Results}
        \begin{columns}[T]
            \begin{column}{10.6 cm}
                \begin{figure}
                    \caption{\parbox[t]{8.6cm}{Monthly electricity generation by source -- Austria, 2030 (a) maximum wind power, (b) maximum solar power}}
                    \includegraphics[width=0.875\textwidth]{./figures/dispatch_AT.pdf} \label{fig:monthly_gen}
                \end{figure}
            \end{column}
            \begin{column}{5.0 cm}
                \bigskip
                \begin{itemize}
                    \item Wind power is favourably complementary to run-of-river generation
                    \item Solar PV positively correlated with run-of-river generation
                    \item In PV-based system more imports needed in winter, while export price in summer lower
                \end{itemize}
            \end{column}
        \end{columns}
    \end{frame}

    \begin{frame}
        \frametitle{Results}
        %\framesubtitle{PV capital cost: \euro 625/kWp}
        \begin{columns}[T]
            \begin{column}{10.6 cm}
                \begin{figure}
                    \caption{\parbox[t]{8.4cm}{Annual opportunity cost of wind power -- Austria, 2030 PV capital cost: \euro 625/kWp}}
                    \includegraphics[width=0.975\textwidth]{./figures/oc_base.pdf} \label{fig:opportunity_cost_base}
                \end{figure}
            \end{column}
            \begin{column}{5.0 cm}
                \bigskip
                \begin{itemize}
                    \item Figure shows on marginal system cost of not erecting wind turbines
                    \item Policy target implies marginal opportunity cost of 22\,500 \euro/MW per year
                    \item Equals \euro 1.3 mn over lifetime of a single standard wind turbine
                \end{itemize}
            \end{column}
        \end{columns}
    \end{frame}

    \begin{frame}
        \frametitle{Results}

        \begin{columns}[T]
            \begin{column}{10.6 cm}
                \begin{figure}
                    \caption{\parbox[t]{8.0cm}{Sensitivity of annual opportunity cost of wind power PV capital cost: \euro 625/kWp}}
                    \includegraphics[width=0.975\textwidth]{./figures/Sensitivities.pdf} \label{fig:sensitivities}
                \end{figure}
            \end{column}
            \begin{column}{5.4 cm}
                \bigskip
                \begin{itemize}
                    \item \emph{Low capital cost of PV} assumes overnight cost of PV of \euro 560/kWp
                    \item \emph{No bottleneck} assumes 10 GW transmission capacity
                \end{itemize}
            \end{column}
        \end{columns}
    \end{frame}

%    \begin{frame}
%        \frametitle{Summary}
%        \begin{itemize}
%            \item So-called 100\% goal necessarily turns Austria into a net exporter of electricity
%            \item Choice of RET impacts installed capacities, system operation, and, most importantly, trade flows
%            \item Austria's wind resources are favourable for exporting to Germany
%            \item Export revenues accrue to electricity producers, not consumers
%            \item Consumers benefit from low electricity prices
%            \item Compensation for neighbors of wind parks makes sense
%        \end{itemize}
%    \end{frame}

    \begin{frame}
        \frametitle{Discussion}
        \begin{itemize}
            \item Renewable resource quality held constant as capacity is expanded
            \item sub-national electricity grids neglected
            \item external cost of open-space PV assumed to be zero
        \end{itemize}
    \end{frame}

    \begin{frame}
        \frametitle{Conclusions}
        \begin{itemize}
            \item If cost of $CO_2$ low and PV is realized as open-space, utility-scale installations (with zero external cost),
            wind power does not come at a system cost advantage
            \item With high $CO_2$ cost, a preference for rooftop PV, or in the presence of negative external effects of
            open-space, utility-scale PV, wind power can reduce total system cost
            \item Wind power induces a current account surplus, which accrues to exporters. Compensation of affected
            residents could improve distributional fairness and, potentially, increase acceptance of wind power
            \item In contrast to (relative) market value, opportunity cost of renewables consistent with fully renewable energy systems
            \item Trade effects lead to much higher system cost impacts of capacity decisions than currently acknowledged in literature
        \end{itemize}
    \end{frame}

    \begin{frame}
        \frametitle{Thank you!}
        %\vspace{1 cm}
        \begin{center}
            \structure{%
                \href{https://arxiv.org/abs/2006.08009}{The Cost of Undisturbed Landscapes}
            }\\%
            \vspace{0.6cm}
            \qrcode[height=2cm]{https://arxiv.org/abs/2006.08009}\\
            \vspace{0.6 cm}
            \href{https://arxiv.org/abs/2006.08009}{https://arxiv.org/abs/2006.08009}
        \end{center}
    \end{frame}
\end{document}
